---
title: "Simulating GRM and Pheno"
author: "BENJAMIN"
date: "2/9/2023"
output: html_document
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library("OmicKriging")
library(tibble)
library(lme4qtl)
library(tidyverse)
library(rio)
library("genio")
library(micEcon)
library("symMatrix")
install.packages("micEcon")
setwd("/Users/bm0211/RegEx/Turkana_stuff")

```

## This Chunk is for simulating a dataset that we already know the answer to, 
## so that we can test our pipeline just to make sure that the pipeline works 
## and that we get the answer we expext given that we already know what the heritability should be

```{r}
##Simulate the Full Sibling Design GRM matrix
simulateKinship <- function(N, L = 5)
{
  if(!require(Matrix))
    stop('Could not load package Matrix')
  
  stopifnot(!(N %% L)) # family size = L
  Nl <- N / L
  
  # kinship of a single family
  K0 <- matrix(0.5, L, L)
  diag(K0) <- 1
  K0[1, 2] <- 0
  K0[2, 1] <- 0
  
  K <- kronecker(Diagonal(Nl), K0) 
  
  ids <- seq(N)
  rownames(K) <- ids
  colnames(K) <- ids
  
  return(K)
}
##################################################
N  = 870 ##870 is the total number of Turkana we have sequenced so far
h2 = 0.5 ##Heritability should be equal to the value set here
sigma = 1
G = simulateKinship(N)
##################################################
##Step 2 generate Breeding values from the GRM, and use them to Create a Phenotype File
add = sigma * sqrt(h2) * t(chol(G)) %*% rnorm(N)
#################################################
##Step 3 Create a Phenotype file using the generated Breeding values in step 2
z = add + rnorm(N, sd = sigma * sqrt(1-h2)) # P = G + E
##################################################
##Step 4, Write the GRM and Phenotype file into a format where I can attach the Unique.IDs 
## from my pipeline
G = as_tibble(as.matrix(G)) # print in correct GRM file format
Z= as.data.frame(as.matrix(z)) # add to correct phenotype file format 
# run h2 estimate using Z and G above after getting them in correct format
```
## Simulating phenotype with real GRM
```{r}
symmetrise <- function(mat)
{rowscols <- which(lower.tri(mat), arr.ind=TRUE)
  sparseMatrix(i=rowscols[,1],
  j=rowscols[,2],
  x=mat[lower.tri(mat)],
  symmetric=TRUE, 
  dims=c(nrow(mat),
  ncol(mat)))
}

N  = 870 ##870 is the total number of Turkana we have sequenced so far
h2 = 0.5 ##Heritability should be equal to the value set here
sigma = 1
G_real = read_GRMBin("Recent_CleanHDL.grm") ##The REAL GRM from the cluster
G_real = as.matrix(G_real)
G_real = Matrix(G_real)
##G_real <- symmetrise(G_real) 
G_real[lower.tri(G_real)] = t(G_real)[lower.tri(G_real)]
G_real <- miscTools::symMatrix(G_real[upper.tri(G_real, TRUE)], nrow=nrow(G_real), byrow=TRUE)
G_real <- forceSymmetric(G_real)
##################################################
##Step 2 generate Breeding values from the GRM, and use them to Create a Phenotype File
add_real = sigma * sqrt(h2) * t(chol(G_real)) %*% rnorm(N)
#################################################
##Step 3 Create a Phenotype file using the generated Breeding values in step 2
z_real = add_real + rnorm(N, sd = sigma * sqrt(1-h2)) # P = G + E
##################################################
##Step 4, Write the GRM and Phenotype file into a format where I can attach the Unique.IDs 
## from my pipeline
G_real = as_tibble(as.matrix(G_real)) # print in correct GRM file format
Z_real= as.data.frame(as.matrix(z_real)) # add to correct phenotype file format 
# run h2 estimate using Z and G above after getting them in correct format

```
## The chunk Below is for Getting the simulated Phenotype and GRM files above into correct format 
## for GCTA
## One option is to Write Phenotype file then edit in excel it to make sure it follows the GCTA format of 
## 6 columns (family_id, individual_id, paternal_id, maternal_id, sex, phenotype) 
## Here, you can add the Barcode and IDs that match our real data
## CODE: write.table(Z, file = "Simulated_Phenotype.csv",row.names = FALSE, quote=FALSE)
## After this, read in the edited Phenotype file in the correct format and write it into 
## a .Phen file which is what GCTA uses for Heritability estimates
```{r}
## Second option to assign the REAL IDs computationally
## First Load the Gene_IDs
Gene_IDS <- import("Gene_IDs.csv") ##Gene IDS as copied from the .fam file on the cluster
Gene_IDS <- Gene_IDS %>% mutate (V1 = recode(V1 , "0"="-9")) %>% mutate (V2 = recode(V2 , "0"="-9"))%>% mutate (V3 = recode(V3 , "0"="-9"))
head(Gene_IDS)
sim_data <- merge(data.frame(Gene_IDS, row.names = NULL), data.frame(Z, row.names = NULL), by = 0, all = TRUE)[-1]
dim(sim_data)
names(sim_data)[7] <- "simulated_values"
head(sim_data)
############################################################################
##Write table to Use on the Cluster with GCTA
write.table(sim_data, file = "sim_data.phen",row.names = FALSE, quote=FALSE)

#######################################################################
##Write the Simulated GRM from this code into the GRM format that GCTA recognizes 
##usually has three file types (.grm.bin, .grm.N.bin, .grm.id)
kinship <- matrix(c(as.matrix(G)), nrow = 870)
M <- matrix(Gene_IDS$BARCODE_ID, byrow = TRUE, nrow = 870, ncol = 870)
dim(M)
head(M)
fam <-tibble(
  fam = Gene_IDS$FAM_ID,
  id = Gene_IDS$BARCODE_ID
)
name <- "HDL_Simulated"
write_grm(name, kinship, M = M, fam = fam, shape = c("square"))
## Use the simulated GRM and Phenotype files with the GCTA package on the cluster
## ....test if you will get the expected Heritability (in this case ~0.5)
```

## The Chunk below is to test the simulated Data with an R package used to check heritability
```{r}
##Simulation should give heritability ~0.5
pheno_df = sim_data
colnames(pheno_df)
names(pheno_df)[7] <- "pheno_simulated"
names(pheno_df)[2] <- "id"
##
##
grm = read_GRMBin("HDL_Simulated.grm", size = 4) ##The simulated GRM from the chunk above

## Run the Package to assess heritability
model = relmatLmer(pheno_simulated ~ (1|id),pheno_df, relmat = list(id = as.matrix(grm)))
lme4qtl::VarProp(model)
##You should get a H2 estimate of ~0.5 if the pipeline worked well (last column "Prop")
##Below is with actual data
```

## The Chuck Below is for running a heritability estimate using our actual data using the R package
## This Chunk is for HDL heritability ...Substitute the .phen file incase you want another trait
```{r}
setwd("/Users/bm0211/RegEx/Turkana_stuff")
##Import the REAL Phenotype file from another script, merge with the simulated data
Phenotype_HDL_raw <- read.table("Pulse_pressure_Unique.phen", header = T) ##HDL .phen file
Phenotype_HDL_raw <- Phenotype_HDL_raw %>% distinct(indiv_id, .keep_all = TRUE)
dim(Phenotype_HDL_raw)
colnames(Phenotype_HDL_raw)
## Create DF with REAL and Simulated data with IDs matching the REAL GRM 
## ...containS the simulated phenotype (pheno_df created above)
Phenotype_HDL <- merge(Phenotype_HDL_raw , pheno_df, by.x = "indiv_id", by.y = "id", all.x = TRUE)
head(Phenotype_HDL)
dim(Phenotype_HDL)
colnames(Phenotype_HDL)
##The merge creates a .x and .y for the pheno column ...
## ....Pheno.x is the REAL data while pheno.y is the simulated data
Phenotype_HDL$pheno <- as.numeric(Phenotype_HDL$pheno)
Phenotype_HDL <- Phenotype_HDL %>% distinct(indiv_id, .keep_all = TRUE)
dim(Phenotype_HDL)
str(Phenotype_HDL)
##Heritability estimates using R
GRM_HDL = read_GRMBin("Recent_CleanHDL.grm", size = 4) ##The REAL GRM from the cluster
GRM_HDL = as.matrix(GRM_HDL)
#colnames(GRM_HDL)
tmp_df = subset(Phenotype_HDL, select = c(2,1,3:15))
dim(tmp_df)
colnames(tmp_df)
head(tmp_df)
## Heritability with Actual Data
HDL_model = relmatLmer(pheno ~ (1|indiv_id),tmp_df, relmat = list(indiv_id = as.matrix(GRM_HDL)))
lme4qtl::VarProp(HDL_model)

##Heritability with simulated Data ...sanity check ..I set my data to have ~0.5 heritability
HDL_model_sim = relmatLmer(pheno_simulated ~ (1|indiv_id),tmp_df, relmat = list(indiv_id = as.matrix(GRM_HDL)))
lme4qtl::VarProp(HDL_model_sim)
##write.table(tmp_df, file = "HDL_Unique_sim_data.phen", row.names = FALSE, quote = FALSE)
```